---
title: '3220103172'
author: "Zhehao Chen"
date: "2024-07-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1.
##### (a) (5,8,1,5,8,1,...,5,8,1) where there are 10 occurrences of 5.
To create a vector that repeats the sequence (5, 8, 1) so that there are 10 occurrences of 5, you can use the rep function:
```{r}
tmp <- c(5, 8, 1)
vec_a <- rep(tmp, times = 10)
print(vec_a)
```
##### (b) (5,5,...,5,8,8,...,8,1,1,...,1) where there are 30 occurrences of 5, 20 occurrences of 8 and 10 occurrences of 1.
To create a vector with 30 occurrences of 5, 20 occurrences of 8, and 10 occurrences of 1, you can use the rep function with the times argument as a vector:

```{r}
tmp <- c(5, 8, 1)
times <- c(30, 20, 10)
vec_b <- rep(tmp, times = times)
print(vec_b)
```



### 2.
Execute the following lines which create two vectors of random integers which are chosen with replacement from the integers 0,1,...,999. Both vectors have length 250.
```{r}
 xVec <- sample(0:999, 250, replace=T)
 yVec <- sample(0:999, 250, replace=T)
 xVec
 yVec
```
##### (a) Create the vector (y2 − x1,··· ,yn − xn−1).
yVec[2:250] - xVec[1:249] creates a vector by subtracting each element in xVec (from the 1st to the 249th) from the corresponding next element in yVec (from the 2nd to the 250th).
```{r}
aVec <- yVec[2:250] - xVec[1:249]
print(aVec)
```
##### (b) Pick out the values in yVec which are > 600.
yVec[yVec > 600] selects the values from yVec that are greater than 600.
```{r}
bVec <- yVec[yVec > 600]
print(bVec)
```
##### (c) What are the index positions in yVec of the values which are > 600?
which(yVec > 600) returns the index positions of elements in yVec that are greater than 600.
```{r}
cVec <- which(yVec > 600)
print(cVec)
```
##### (d) Sort the numbers in the vector xVec in the order of increasing values in yVec.
xVec[order(yVec)] sorts xVec based on the order of the corresponding elements in yVec.
```{r}
dVec <- xVec[order(yVec)]
print(dVec)
```
#####  (e) Pick out the elements in yVec at index positions 1,4,7,10,13,···
yVec[seq(1, length(yVec), by = 3)] picks out every third element in yVec starting from the first position.
```{r}
eVec <- yVec[seq(1, length(yVec), by = 3)]
print(eVec)
```



### 3.
```{r}
X <- c(34, 33, 65, 37, 89, NA, 43, NA, 11, NA, 23, NA)
X
```
##### Write a piece of R code to count the number of occurrences of NA in X?
 is.na(X) returns a logical vector of the same length as X with TRUE for each NA and FALSE otherwise.\
 sum(is.na(X)) sums the TRUE values, effectively counting the number of NA values in X.
```{r}
na_count <- sum(is.na(X))
print(na_count)
```



### 4.
For this problem we’ll use the (built-in) dataset state.x77.
```{r}
library(tidyverse)
data(state)
state.x77 <- as_tibble(state.x77, rownames = 'State')
```
##### (a) Select all the states having an income less than 4300, and calculate the average income of these states.
```{r}
# Select states with income less than 4300
low_income_states <- state.x77 %>% filter(Income < 4300)

# Calculate the average income of these states
average_low_income <- mean(low_income_states$Income)
average_low_income
```
##### (b) Sort the data by income and select the state with the highest income.
```{r}
# Sort the data by income
sorted_states <- state.x77 %>% arrange(desc(Income))

# Select the state with the highest income
highest_income_state <- sorted_states %>% slice(1)
highest_income_state
```
##### (c) Add a variable to the data frame which categorizes the size of population: <= 4500 is S, $> 4500 $ is L.
```{r}
# Add a variable categorizing the size of population
state.x77 <- state.x77 %>% 
  mutate(PopulationSize = ifelse(Population <= 4500, 'S', 'L'))
head(state.x77)
```
##### (d) Find out the average income and illiteracy of the two groups of states, distinguishing by whether the states are small or large.
```{r}
# Find the average income and illiteracy of the two groups of states
grouped_stats <- state.x77 %>% 
  group_by(PopulationSize) %>% 
  summarise(
    Average_Income = mean(Income),
    Average_Illiteracy = mean(Illiteracy)
  )

grouped_stats
```



### 5.
#####  (a) Write a function to simulate n observations of (X1,X2) which follow the uniform distribution over the square [0,1] × [0,1].
simulate_observations(n) generates n pairs of (X1,X2) uniformly distributed over [0,1]×[0,1].
```{r}
simulate_observations <- function(n) {
  X1 <- runif(n, 0, 1)
  X2 <- runif(n, 0, 1)
  return(data.frame(X1, X2))
}
```
#####  (b) Write a function to calculate the proportion of the observations that the distance between (X1,X2) and the nearest edge is less than 0.25, and the proportion of them with the distance to the nearest vertex less than 0.25.
edge_proportion: The proportion of observations where the distance to the nearest edge is less than 0.25.\
vertex_proportion: The proportion of observations where the distance to the nearest vertex is less than 0.25.
```{r}
calculate_proportions <- function(observations) {
  n <- nrow(observations)
  X1 <- observations$X1
  X2 <- observations$X2
  
  # Distance to the nearest edge
  edge_distances <- pmin(X1, 1 - X1, X2, 1 - X2)
  edge_proportion <- mean(edge_distances < 0.25)
  
  # Distance to the nearest vertex
  vertex_distances <- pmin(
    sqrt((X1 - 0)^2 + (X2 - 0)^2),  # distance to (0,0)
    sqrt((X1 - 0)^2 + (X2 - 1)^2),  # distance to (0,1)
    sqrt((X1 - 1)^2 + (X2 - 0)^2),  # distance to (1,0)
    sqrt((X1 - 1)^2 + (X2 - 1)^2)   # distance to (1,1)
  )
  vertex_proportion <- mean(vertex_distances < 0.25)
  
  return(list(edge_proportion = edge_proportion, vertex_proportion = vertex_proportion))
}
```
##### Example Usage:
Here's how you can use the functions to simulate observations and calculate the proportions:
```{r}
# Simulate 1000 observations
set.seed(123)  # Setting seed for reproducibility
observations <- simulate_observations(1000)

# Calculate proportions
proportions <- calculate_proportions(observations)
print(proportions)
```



### 6.
Mortality rates per 100,000 from male suicides for a number of age groups and a number of countries are given in the following data frame.
```{r}
 library(tidyverse)
 suicrates <- tibble(Country = c('Canada', 'Israel', 'Japan', 'Austria', 'France', 'Germany',
 'Hungary', 'Italy', 'Netherlands', 'Poland', 'Spain', 'Sweden', 'Switzerland', 'UK', 'USA'),
 Age25.34 = c(22, 9, 22, 29, 16, 28, 48, 7, 8, 26, 4, 28, 22, 10, 20),
 Age35.44 = c(27, 19, 19, 40, 25, 35, 65, 8, 11, 29, 7, 41, 34, 13, 22),
 Age45.54 = c(31, 10, 21, 52, 36, 41, 84, 11, 18, 36, 10, 46, 41, 15, 28),
 Age55.64 = c(34, 14, 31, 53, 47, 49, 81, 18, 20, 32, 16, 51, 50, 17, 33),
 Age65.74 = c(24, 27, 49, 69, 56, 52, 107, 27, 28, 28, 22, 35, 51, 22, 37))
```
##### (a) Transform suicrates into long form.
```{r}
 library(tidyr)

# Transforming the data to long form
suicrates_long <- suicrates %>%
  pivot_longer(cols = starts_with("Age"), names_to = "AgeGroup", values_to = "Rate")

print(suicrates_long)
```
##### (b) Construct side-by-side box plots for the data from different age groups, and comment on what the graphic tells us about the data.
```{r}
 library(ggplot2)

# Constructing side-by-side box plots
ggplot(suicrates_long, aes(x = AgeGroup, y = Rate)) +
  geom_boxplot() +
  labs(title = "Suicide Rates by Age Group",
       x = "Age Group",
       y = "Suicide Rate per 100,000") +
  theme_minimal()
```



### 7.
 In the data set pressure, the relevant theory is that associated with the Claudius- Clapeyron equation, by
 which the logarithm of the vapor pressure is approximately inversely proportional to the absolute temperature
 (temperature + 273). Transform the data in the manner suggested by this theoretical relationship, plot
 the data, fit a regression line, and add the line to the graph. Does the fit seem adequate?
##### (1) Load the 'pressure' data set
```{r}
 data(pressure)
head(pressure)
```
##### (2) Transform the data
According to the Clausius-Clapeyron equation, the logarithm of the vapor pressure is approximately inversely proportional to the absolute temperature (temperature + 273). We'll transform the temperature to its absolute value and compute the logarithm of the vapor pressure:
```{r}
pressure$TemperatureK <- pressure$temperature + 273
pressure$log_pressure <- log(pressure$pressure)
```
##### (3) Plot the transformed data
```{r}
plot(1 / pressure$TemperatureK, pressure$log_pressure,
     xlab = "1 / Temperature (1/K)",
     ylab = "Log(Pressure)",
     main = "Clausius-Clapeyron Plot")
```
##### (4) Fit a regression line
Fit a linear model to the transformed data:
```{r}
fit <- lm(log_pressure ~ I(1 / TemperatureK), data = pressure)
summary(fit)
```
##### (5) Add the regression line to the plot
##### Here's the complete R code for the entire process:
```{r}
# Load the pressure data set
data(pressure)
head(pressure)

# Transform the data
pressure$TemperatureK <- pressure$temperature + 273
pressure$log_pressure <- log(pressure$pressure)

# Plot the transformed data
plot(1 / pressure$TemperatureK, pressure$log_pressure,
     xlab = "1 / Temperature (1/K)",
     ylab = "Log(Pressure)",
     main = "Clausius-Clapeyron Plot")

# Fit a regression line
fit <- lm(log_pressure ~ I(1 / TemperatureK), data = pressure)
summary(fit)

# Add the regression line to the plot
abline(fit, col = "red")
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
