---
title: "Ch6"
author: "CZH"
date: "2024-07-06"
output:
  pdf_document: default
  html_document: default
---

## Test Case: Birth weight data
Included in R already:
```{r}
library(tidyverse)
library(lubridate)
library(MASS)
library(conflicted)
data(birthwt)
summary(birthwt)
```

#### Make it readable!
```{r}
colnames(birthwt)
```
```{r}
colnames(birthwt) <- c("birthwt.below.2500", "mother.age",
                       "mother.weight", "race",
                       "mother.smokes", "previous.prem.labor",
                       "hypertension", "uterine.irr",
                       "physician.visits", "birthwt.grams")
```
Let's make all the factors more descriptive.
```{r}
birthwt$race <- factor(c("white", "black", "other")[birthwt$race])
birthwt$mother.smokes <- factor(c("No", "Yes")[birthwt$mother.smokes + 1])
birthwt$uterine.irr <- factor(c("No", "Yes")[birthwt$uterine.irr + 1])
birthwt$hypertension <- factor(c("No", "Yes")[birthwt$hypertension + 1])
```

#### Bar plot for race
```{r}
birthwt |> ggplot(aes(x = race))+
  geom_bar()+
  labs(title = "Count of Mother's Race in Springfield MA, 1986")
```
#### Scatter plot for mother's ages
```{r}
birthwt |> ggplot(aes(x = 1:nrow(birthwt), y = mother.age))+
  geom_point()+
  labs(x = 'number', title = "Mother's Ages in Springfield MA, 1986")
```
#### Sorted mother's ages
```{r}
birthwt |> arrange(mother.age) |> ggplot(aes(x = 1:nrow(birthwt), y = mother.age))+
  geom_point()+
  labs(x = 'number', title = "Mother's Ages in Springfield MA, 1986")
```
#### Birth weight versus mother's ages
```{r}
birthwt |> ggplot(aes(x = mother.age, y = birthwt.grams))+
  geom_point()+
  labs(title = "Birth Weight by Mother's Age in Springfield MA, 1986")
```
#### Boxplot
Let's fit some models to the data pertaining to our outcome(s) of interest.
```{r}
birthwt |> ggplot(aes(x = mother.smokes, y = birthwt.grams))+  
  geom_boxplot()+
  labs(title = "Birth Weight by Mother's Smoking Habit", y = "Birth Weight (g)", x="Mother Smokes")
```

#### Basic statistical testing
```{r}
t.test (birthwt$birthwt.grams[birthwt$mother.smokes == "Yes"], 
        birthwt$birthwt.grams[birthwt$mother.smokes == "No"], var.equal = T)
```
Does this difference match the linear model?
```{r}
linear.model.1 <- lm (birthwt.grams ~ mother.smokes, data=birthwt)
summary(linear.model.1)
```
```{r}
linear.model.2 <- lm (birthwt.grams ~ mother.age, data=birthwt)
summary(linear.model.2)
```
```{r}
plot(linear.model.1)
plot(linear.model.2)
```


#### Detecting Outliers
These are the default diagnostic plots for the analysis. Note that our oldest mother and her heaviest child are greatly skewing this analysis as we suspected.
```{r}
birthwt.noout <- birthwt |> dplyr::filter(mother.age <= 40)
linear.model.3 <- lm (birthwt.grams ~ mother.age, data=birthwt.noout)
summary(linear.model.3)
```


#### More complex models
Add in smoking behavior:
```{r}
linear.model.3a <- lm (birthwt.grams ~ + mother.smokes + mother.age, data=birthwt.noout)
summary(linear.model.3a)
```
```{r}
plot(linear.model.3a)
```

```{r}
linear.model.3b <- lm (birthwt.grams ~ mother.age + mother.smokes + race, data=birthwt.noout)
summary(linear.model.3b)
```
```{r}
plot(linear.model.3b)
```

#### Everything Must Go (In)
```{r}
linear.model.4 <- lm (birthwt.grams ~ ., data=birthwt.noout)
summary(linear.model.4)
```

#### Everything Must Go (In), Except What Must Not
One of those variables was birthwt.below.2500 which is a function of the outcome.
```{r}
linear.model.4a <- lm (birthwt.grams ~ . - birthwt.below.2500, data=birthwt.noout)
summary(linear.model.4a)
```
```{r}
plot(linear.model.4a)
```

#### Generalized Linear Models
Maybe a linear increase in birth weight is less important than if it's below a threshold like 2500 grams (5.5 pounds). Let's fit a generalized linear model instead:
```{r}
glm.0 <- glm (birthwt.below.2500 ~ . - birthwt.grams, data=birthwt.noout)
summary(glm.0)
```
```{r}
plot(glm.0)
```

The default value is a Gaussian model (a standard linear model). Change this:
```{r}
glm.1 <- glm (birthwt.below.2500 ~ . - birthwt.grams, data=birthwt.noout, family=binomial(link=logit))
summary(glm.1)
```
```{r}
plot(glm.1)
```

#### What Do We Do With This, Anyway?
Let's take a subset of this data to do predictions.
```{r}
odds <- seq(1, nrow(birthwt.noout), by=2)
birthwt.in <- birthwt.noout[odds,]
birthwt.out <- birthwt.noout[-odds,]
linear.model.half <- lm (birthwt.grams ~ . - birthwt.below.2500, data=birthwt.in)
summary (linear.model.half)
```
```{r}
birthwt.predict <- predict (linear.model.half)
cor (birthwt.in$birthwt.grams, birthwt.predict)
```
```{r}
tibble(x = birthwt.out$birthwt.grams, y = birthwt.predict) |>
  ggplot (aes(x = x, y = y)) + geom_point()
```
```{r}
birthwt.predict.out <- predict (linear.model.half, birthwt.out)
cor (birthwt.out$birthwt.grams, birthwt.predict.out)
```
```{r}
tibble(x = birthwt.out$birthwt.grams, y = birthwt.predict.out) |>
  ggplot (aes(x = x, y = y)) + geom_point()
```

## Random number generators
Example: runif, where we know where it started
```{r}
runif(1:10)
```
```{r}
set.seed(10)
runif(1:10)
```

#### Basic version: Linear Congruential Generator
```{r}
seed <- 10
new.random <- function (a=5, c=12, m=16) {
  out <- (a*seed + c) %% m
  seed <<- out
  return(out)
}
out.length <- 20
variates <- rep (NA, out.length)
for (kk in 1:out.length) variates[kk] <- new.random()
variates
```
Try again. Period 8:
```{r}
variates <- rep (NA, out.length)
for (kk in 1:out.length) variates[kk] <- new.random(a=131, c=7, m=16)
variates
```
Try again, again. Period 16:
```{r}
variates <- rep (NA, out.length)
for (kk in 1:out.length) variates[kk] <- new.random(a=129, c=7, m=16)
variates
```
Try again, at last. Numerical Recipes uses:
```{r}
variates <- rep (NA, out.length)
for (kk in 1:out.length) variates[kk] <- new.random(a=1664545, c=1013904223, m=2^32)
variates
```

## A few distributions of interest:
Uniform(0,1)\
Bernoulli(p)\
Binomial(n,p)\
Gaussian(0,1)\
Exponential(1)\
Gamma(a)\

## Suppose we were working with the Exponential distribution.
#### rexp() generates variates from the distribution.
```{r}
rexp(0:5)
```

#### dexp() gives the probability density function.
```{r}
dexp(0:5)
```
```{r}
this.range <- 0:50/5
plot (this.range, dexp(this.range), ty="l")
lines (this.range, dexp(this.range, rate=0.5), col="red")
lines (this.range, dexp(this.range, rate=0.2), col="blue")
```
#### pexp() gives the cumulative distribution function.
```{r}
pexp(0:5)
```
```{r}
this.range <- 0:50/5
plot (this.range, pexp(this.range), ty="l")
lines (this.range, pexp(this.range, rate=0.5), col="red")
lines (this.range, pexp(this.range, rate=0.2), col="blue")
```

#### qexp() gives the quantiles.
```{r}
qexp(0:5)
```
```{r}
this.range <- seq(0,1,by=0.01)
plot (this.range, qexp(this.range), ylim = c(0, 10), ty="l")
lines (this.range, qexp(this.range, rate=0.5), col="red")
lines (this.range, qexp(this.range, rate=0.2), col="blue")
```

## S\&P 500
The Standard and Poor's 500, or simply the S\&P 500, is a stock market index tracking the stock performance of 500 large companies listed on exchanges in the United States. It is one of the most commonly followed equity indices.
```{r}
library(readxl)
SP <- read_excel("data/Stock_Bond.xls") |> dplyr::select(Date, `S&P_AC`) |>
  rename(Index = `S&P_AC`)
SP |> ggplot(aes(x = Date, y = Index)) + geom_line()
```
\
The price $p_t$ doesn't matter, what matters are the returns $r_t=log(p_t/p_{t-1})$
```{r}
returns <- na.omit(as.vector(diff(log(SP$Index))))
summary(returns)
```
```{r}
plot(returns, type="l")
```
## The Data's Distribution
#### quantile(x,probs) calculates the quantiles at probs from x.
```{r}
quantile(returns,c(0.25,0.5,0.75))
```
#### ecdf() - e mpirical c umulative d istribution f unction; no assumptions but also no guess about distribution between the observations.
```{r}
plot(ecdf(returns), main="Empirical CDF of S&P 500 index returns")
```
Conceptually, quantile and ecdf are inverses to each other.

## Getting Probability Densities from Data
ist(x) calculates a histogram from x.\
divide the data range up into equal-width bins and count how many fall into each bin.\
Or divide bin counts by (total count)*(width of bin), and get an estimate of the probability density function (pdf).\
```{r}
hist(returns,n=101,probability=TRUE)
```
density(x) estimates the density of x by counting how many observations fall in a little window around each point, and then smoothing.\
"Bandwidth" $[&approx&]$ width of window around each point\
Technically, a "kernel density estimate"\
Remember: density() is an estimate of the pdf, not The Truth
density returns a collection of $x,y$ values, suitable for plotting
```{r}
plot(density(returns),main="Estimated pdf of S&P 500 index  returns")
```
```{r}
hist(returns,n=101,probability=TRUE)
lines(density(returns),lty="dashed")
```
table() - tabulate outcomes, most useful for discrete spaces; remember to normalize if you want probabilities.
```{r}
plot(table(cats$Sex)/nrow(cats),ylab="probability")
```
## R commands for distributions
dfoo = the probability d ensity (if continuous) or probability mass function of foo (pdf or pmf)\
pfoo = the cumulative p robability function (CDF)\
qfoo = the q uantile function (inverse to CDF)\
rfoo = draw r andom numbers from foo (first argument always the number of draws)\
?Distributions to see which distributions are built in.

## Displaying Probability Distributions
curve is very useful for the d, p, q functions:
```{r}
curve(dgamma(x,shape=45,scale=1.9),from=0,to=200)
```




